ProteInfer: deep networks for protein functional inference

### 2021-09-20 latex/draft3.tex
Updates

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for predicting functional properties of proteins}
\shorttitle{ProteInfer}
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[*,1,\Letter]{Theo Sanderson}
\author[*,2]{Maxwell Bileschi}
\author[2]{David Belanger}
\author[2]{Lucy Colwell}

@@

\begin{abstract}
Predicting the function(s) of a protein from its amino acid sequence has been a long-standing challenge for the bioinformatics community. Standard approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. In our work, we instead employ deep convolutional neural networks to directly predict a variety of protein functions - EC numbers and GO terms - directly from an unaligned amino acid sequence. This performs comparably to alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces. In response, we have provided an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. To read the interactive version of this paper, please visit \url{https://google-research.github.io/proteinfer/}.
\end {abstract}

@@

By analogy to deep networks for images, the early layers in these models build up an understanding of simple features such as edges, and later layers use these features to identify textures, and then entire objects. Edge detecting filters can thus be trained with information from all the labelled examples, and the same filters can be used to detect, for instance, both oranges and lemons \cite{activationatlas}.

In response, recent work has contributed a number of deep neural network models for protein function classification\citep{ecpred,deepgo,prolango,deeploc,dspace,deepred,deepre,deepsf,littmann2021embeddings}. These approaches train a single model to recognise multiple properties, building representations of different protein functions via a series of layers, which allow the same low-level features to be used for different high-level classifications. Of special note is the layer preceding the final layer of the network, which constructs an ``embedding'' of the entire example in a high-dimensional vector space, and often captures semantic features of the input.

@@

To address this challenge we employ deep dilated convolutional networks to learn the mapping between full-length protein sequences and functional annotations. The resulting ProteInfer models take amino acid sequences as input and are trained on the well-curated portion of the protein universe annotated by Swiss-Prot \cite{uniprotlatest}. We find that: 1) ProteInfer models reproduce curator decisions for a  variety of functional properties across sequences distant from the training data, 2) attribution analysis shows that the predictions are driven by relevant regions of each protein sequence, and 3) ProteInfer models create a generalised mapping between sequence space and the space of protein functions, which is useful for tasks other than those for which the models were trained. We provide trained ProteInfer networks that enable other researchers to reproduce the analysis presented and to explore embeddings of their proteins of interest, via both a command line tool\footnote{\url{https://github.com/google-research/proteinfer}}, and also via an in-browser JavaScript implementation that demonstrates the computational efficiency of deep-learning approaches.
\section*{Methods}
\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{Figures/figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold; varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

Each model was trained for about 60 hours using the Adam optimizer \citep{kingma2014adam} on 8 NVIDIA P100 GPUs with data parallelism \cite{dean2012large, dataparalellism}.  We found that using more than one GPU for training improved training time by allowing an increased batch size, but did not have a substantial impact on accuracy compared to training for longer with a smaller learning rate and smaller batch size on one GPU. The models have a small set of hyperparameters, such as the number of layers and the number of filters in each layer, which were tuned using random sampling to maximize $\texttt{F}_{max}$ on the random train-test split. Hyperparameter values are available in the \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">Supplement</a>
.
\subsection*{A machine-learning compatible dataset for protein function prediction}
The UniProt database is the central global repository for information about proteins. The manually curated portion, Swiss-Prot, is constructed by assessing $~$60,000 papers each year to harvest $\gt$ 35\% of the theoretically curatable information in the literature \cite{uniprotlatest}. We focus on Swiss-Prot to ensure that our models learn from human-curated labels, rather than labels generated by a computational annotation pipeline. Each protein in Swiss-Prot goes through a 6-stage process of sequence curation, sequence analysis, literature curation, family-based curation, evidence attribution, and quality assurance.
Functional annotation is stored in UniProt largely through \textit{database cross-references}, which link a specific protein with a label from a particular ontology. These cross-references include: Enzyme Commission (EC) numbers, representing the function of an enzyme; Gene Ontology (GO) terms relating to the protein's molecular function, biological process, or subcellular localisation; protein family information contained in the Pfam \citep{pfamlatest}, SUPFAM\citep{supfamlatest}, PRINTS \citep{prints}, TIGR \citep{tigr}, PANTHR \citep{panther} databases, or the umbrella database InterPro \citep{interpro}; as well as other information including ortholog databases and references on PubMed. Here, we focus on EC and GO labels, though our model training framework can immediately extend to other label sets.

We use two methods to split data into training and evaluation sets. First, a random split of the data allows us to answer the following question: suppose that curators had randomly annotated only 80\% of the sequences in Swiss-Prot. How accurately can ProteInfer annotate the remaining 20\%?  Second, we use UniRef50 \citep{suzek2015uniref} clustering to split the data, to model a challenging use-case in which an unseen sequence has low sequence similarity to anything that has been previously annotated. Note there are further methods for splitting, such as reserving the most recently-annotated proteins for evaluating models. This approach, which is used in CAFA and CASP \citep{zhou2019cafa, gillis}, helps ensure a fair competition because labels for the evaluation data are not available to participants, or the scientific community at large, until after the competition submissions are due. A drawback of this approach is that it is not available for EC numbers, test sets are often small and have potential biases (e.g., they are limited to a small number of species or functions). Further, there is no such dataset for EC classification, which is the primary focus of our analyses below. Finally, note that all of the above approaches typically lack reliable annotation for true negatives\citep{warwick2020benchmarking}.

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}
Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">Supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

@@

We then assessed the ability of this method to more generally localize function within a sequence, even though the model was not trained with any explicit localization information. We selected all enzymes from Swiss-Prot that had two separate leaf-node EC labels for which our model predicted known EC labels, and these labels were mappable to corresponding Pfam labels. For each of these proteins, we obtained coarse-grained functional localization by using CAM to predict the order of the domains in the sequence and compared to the ground truth Pfam domain ordering\textnormal{(see Supplement for details of the method)}%htmlver:(see <a href="./latex/supplement.pdf">Supplement</a> for details of the method)
. We found that in 296 of 304 (97\%) of the cases, we correctly predicted the ordering, though we note that the set of bifunctional enzymes for which this analysis is applicable is limited in its functional diversity \textnormal{(see Supplement)}%htmlver:(see <a href="./latex/supplement.pdf">Supplement</a>)
. Although we did not find that fine-grained, per-residue functional localization arose from our application of CAM, we found that it reliably provided coarse-grained annotation of domains' order, as supported by Pfam. This experiment suggests that this is a promising future area for research.
\subsection*{Neural networks learn a general-purpose embedding space for protein function}
Whereas InterProScan compares each sequence against more than 50,000 %from summing integrated signatures at https://www.ebi.ac.uk/interpro/release_notes.html
individual signatures and BLAST compares against an even larger sequence database, ProteInfer uses a single deep model to extract features from sequences that directly predict protein function. One convenient property of this approach is that in the penultimate layer of the network each protein is expressed as a single point in a high-dimensional space. To investigate to what extent this space is useful in examining enzymatic function, we used the ProteInfer EC model trained on the random split to embed each test set protein sequence into a 1100-dimensional vector.

@@

There is arguably room for improvement in this regard from classical approaches. For example, the online interface to InterProScan can take ~147 seconds to process a 1500 amino acid sequence\footnote{Protein Q77Z83.}\footnote{It should be noted that the individual databases that make up InterProScan may return matches faster, with the online interface to Pfam taking 14--20 seconds for a 1500 amino acid sequence.}, while running the tool may make the search faster, doing so requires downloading a 9 GB file, with an additional 14 GB for the full set of signatures. When uncompressed this installation exceeds 51 GB. Meanwhile, conducting a BLAST search against Swiss-Prot takes 34 seconds for a 1500 amino acid sequence \footnote{A target database with decreased redundancy could be built to reduce this search time, and other optimizations of BLAST have been developed.}.

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see <a href="./latex/supplement.pdf">Supplement</a>
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}
\begin{figure*}[b]
\centering
  \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
  % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
  \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
  \label{fig:cam}
\end{figure*}

@@

We note that there has been extensive work in GO label prediction evaluated on a temporally-split dataset (constructing a test set with the most recently experimentally annotated proteins), e.g., \citep{zhou2019cafa}, and stress that our comparison is instead on random and clustered splits of Swiss-Prot.  As discussed above, our approaches to splitting the data into train and test has different advantages and disadvantages as compared to a temporal split.
%%%GO_METHOD%%%
We trained a single model to predict presence or absence for each of these terms and found that our network was able to achieve a precision of 0.918 and a recall of 0.854 for an F1 score of 0.885 (95\% CI: 0.882--0.887).

@@

There are multiple caveats to these comparisons, both in terms of the trustworthiness of the annotations and how to implement the baselines. One challenge is that the completeness of Swiss-Prot's GO term annotations varies \cite{jiang2014impact}. As an extreme example, \textit{Pan paniscus} (Pygmy Chimpanzee) and \textit{Pan troglodytes} (Chimpanzee) have an identical Apolipoprotein A-II protein,\footnote{Accessions P0DM95 and Q8MIQ5.}, where the first protein has 24 GO annotations, while the latter has 143 GO annotations.\footnote{This count is done using not only the set of all labels that appear in Swiss-Prot, but also any parents of those labels.} One way this is reflected in the performance of the models is that some BLAST matches that have extremely large bit-scores are not annotated identically, and thus hurt the precision of the BLAST model. It is also important to note that our model has the advantage of being specifically trained on the UniProt labelling schema upon which it is being evaluated. InterPro works quite differently, with GO terms being assigned to families, and so inconsistencies in terms of how these are assigned can explain reduced performance -- for instance InterPro families simply do not feature all of the GO terms found in UniProt. Thus these results should be seen as specific to the task of reproducing the curated results in UniProt.

@@

%startleaveout
\section*{Bibliography}
\bibliography{../bibliography}
%\onecolumn
%\newpage
%stopleaveout
%startleaveout

>>>>

%%%TOPSTUFF%%%
\title{ProteInfer: deep networks for protein functional inference}
\shorttitle{ProteInfer}
% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[*,1,\Letter]{Theo Sanderson}
\author[*,2]{Maxwell Bileschi}
\author[2]{David Belanger}
\author[2]{Lucy Colwell}

@@

\begin{abstract}
Predicting the function of a protein from its amino acid sequence has been a long-standing challenge for the bioinformatics community. Standard approaches use sequence alignment to compare a query sequence either to thousands of models of protein families or to large databases of individual protein sequences. In our work, we instead employ deep convolutional neural networks to directly predict a variety of protein functions - EC numbers and GO terms - directly from an unaligned amino acid sequence. This approach provides precise predictions which complement alignment-based methods, and the computational efficiency of a single neural network permits novel and lightweight software interfaces. In response, we have provided an in-browser graphical interface for protein function prediction in which all computation is performed on the user's personal computer with no data uploaded to remote servers. Moreover, these models place full-length amino acid sequences into a generalised functional space, facilitating downstream analysis and interpretation. To read the interactive version of this paper, please visit \url{https://google-research.github.io/proteinfer/}.
\end {abstract}

@@

Deep neural networks have recently transformed a number of labelling tasks, including image recognition -- the early layers in these models build up an understanding of simple features such as edges, and later layers use these features to identify textures, and then entire objects. Edge detecting filters can thus be trained with information from all the labelled examples, and the same filters can be used to detect, for instance, both oranges and lemons \cite{activationatlas}.

In response, recent work has contributed a number of deep neural network models for protein function classification \citep{ecpred,deepgo,prolango,deeploc,dspace,deepred,deepre,deepsf,littmann2021embeddings}. These approaches train a single model to recognise multiple properties, building representations of different protein functions via a series of layers, which allow the same low-level features to be used for different high-level classifications. Of special note is the layer preceding the final layer of the network, which constructs an ``embedding'' of the entire example in a high-dimensional vector space, and often captures semantic features of the input.

@@

To address this challenge we employ deep dilated convolutional networks to learn the mapping between full-length protein sequences and functional annotations. The resulting ProteInfer models take amino acid sequences as input and are trained on the well-curated portion of the protein universe annotated by Swiss-Prot \cite{uniprotlatest}. We find that: 1) ProteInfer models reproduce curator decisions for a  variety of functional properties across sequences distant from the training data, 2) attribution analysis shows that the predictions are driven by relevant regions of each protein sequence, and 3) ProteInfer models create a generalised mapping between sequence space and the space of protein functions, which is useful for tasks other than those for which the models were trained. We provide trained ProteInfer networks that enable other researchers to reproduce the analysis presented and to explore embeddings of their proteins of interest, via both a command line tool\footnote{\url{https://github.com/google-research/proteinfer}}, and also via an in-browser JavaScript implementation that demonstrates the computational efficiency of deep-learning approaches.
\section*{Methods}
\begin{figure}[h!]
\centering
  \includegraphics[width=\columnwidth]{Figures/figure.pdf}%
  \caption{\textbf{A deep dilated convolutional architecture for protein function prediction.} Amino acids are one-hot encoded, then pass through a series of convolutions implemented within residual blocks. Successive filters are increasingly dilated, allowing the top residual layer of the network to build up a representation of high-order protein features. The positional embeddings in this layer are collapsed by mean-pooling to a single embedding of the entire sequence, which is converted into probabilities of each functional classification through a fully connected layer with sigmoidal activations. }
  \label{architecture}
\end{figure}
\subsection*{A neural-network for protein function prediction}
%%%MODEL%%%
In a ProteInfer neural network (\Cref{architecture}%htmlver:<a href="#fig3" target="_self">Figure 3</a>
), a raw amino acid sequence is first represented numerically as a \textit{one-hot} matrix and then passed through a series of convolutional layers. Each layer takes the representation of the sequence in the previous layer and applies a number of \textit{filters}, which detect patterns of features. We use \textit{residual} layers, in which the output of each layer is added to its input to ease the training of deeper networks \cite{resnet}, and dilated convolutions \cite{dilatedconvolutions}, meaning that successive layers examine larger sub-sequences of the input sequence. After building up an embedding of each position in the sequence, the model collapses these down to a single $n$-dimensional embedding of the sequence using average pooling. Since natural protein sequences can vary in length by at least three orders of magnitude, this pooling is advantageous because it allows our model to accommodate sequences of arbitrary length without imposing restrictive modeling assumptions or computational burdens that scale with sequence length. Finally, a fully-connected layer maps these embeddings to logits for each potential label, which are the input to an element-wise sigmoid layer that outputs per-label probabilities. 
We select all labels with predicted probability above a given confidence threshold, and varying this threshold yields a tradeoff between precision and recall. To summarize model performance as a single scalar, we compute the $\texttt{F}_{max}$ score, the maximum $\texttt{F}_{1}$ score (the geometric mean of precision and recall) across all thresholds \citep{radivojac2013large}.

Each model was trained for about 60 hours using the Adam optimizer \citep{kingma2014adam} on 8 NVIDIA P100 GPUs with data parallelism \cite{dean2012large, dataparalellism}.  We found that using more than one GPU for training improved training time by allowing an increased batch size, but did not have a substantial impact on accuracy compared to training for longer with a smaller learning rate and smaller batch size on one GPU. The models have a small set of hyperparameters, such as the number of layers and the number of filters in each layer, which were tuned using random sampling to maximize $\texttt{F}_{max}$ on the random train-test split. Hyperparameter values are available in the \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
.
\subsection*{A machine-learning compatible dataset for protein function prediction}
The UniProt database is the central global repository for information about proteins. The manually curated portion, Swiss-Prot, is constructed by assessing $~$60,000 papers each year to harvest $\gt$ 35\% of the theoretically curatable information in the literature \cite{uniprotlatest}. We focus on Swiss-Prot to ensure that our models learn from human-curated labels, rather than labels generated by a computational annotation pipeline. Each protein in Swiss-Prot goes through a 6-stage process of sequence curation, sequence analysis, literature curation, family-based curation, evidence attribution, and quality assurance.
Functional annotation is stored in UniProt largely through \textit{database cross-references}, which link a specific protein with a label from a particular ontology. These cross-references include: Enzyme Commission (EC) numbers, representing the function of an enzyme; Gene Ontology (GO) terms relating to the protein's molecular function, biological process, or subcellular localisation; protein family information contained in the Pfam \citep{pfamlatest}, SUPFAM\citep{supfamlatest}, PRINTS \citep{prints}, TIGR \citep{tigr}, PANTHR \citep{panther} databases, or the umbrella database InterPro \citep{interpro}; as well as other information including ortholog databases and references on PubMed. Here, we focus on EC and GO labels, though our model training framework can immediately extend to other label sets.

We use two methods to split data into training and evaluation sets. First, a random split of the data allows us to answer the following question: suppose that curators had randomly annotated only 80\% of the sequences in Swiss-Prot. How accurately can ProteInfer annotate the remaining 20\%?  Second, we use UniRef50 \citep{suzek2015uniref} clustering to split the data, to model a challenging use-case in which an unseen sequence has low sequence similarity to anything that has been previously annotated. Note there are alternative methods for splitting, such as reserving the most recently-annotated proteins for evaluating models. This approach, which is used in CAFA and CASP \citep{zhou2019cafa, gillis}, helps ensure a fair competition because labels for the evaluation data are not available to participants, or the scientific community at large, until after the competition submissions are due. Such a split is not available for EC classification, which is the primary focus of our analyses below. Finally, note that all of the above approaches typically lack reliable annotation for true negatives \citep{warwick2020benchmarking}.

To facilitate further development of machine learning methods, we provide TensorFlow \citep{abadi2016tensorflow} TFrecord files for Swiss-Prot \footnote{\url{https://console.cloud.google.com/storage/browser/brain-genomics-public/research/proteins/proteinfer/datasets/}}. Each example has three fields: the UniProt accession, the amino acid sequence, and a list of database cross-reference labels. UniProt annotations include only leaf nodes for hierarchical taxononomies such as EC and GO. To allow machine-learning algorithms to model this hierarchy, we added all parental annotations to each leaf node during dataset creation. 
\section*{Results}
\subsection*{Prediction of catalysed reactions}
We initially trained a model to predict enzymatic catalytic activities from amino acid sequence. This data is recorded as Enzyme Commission (EC) numbers, which describe a hierarchy of catalytic functions. For instance, \textbeta{} amylase enzymes have an EC number of EC:3.2.1.2, which represents the leaf node in the following hierarchy:
%%%ECHIERARCHY%%%
%startleaveout
\renewcommand\DTstyle{\rmfamily}
\setlength{\DTbaselineskip}{15pt}
\medskip
\dirtree{%
.1 {\textbf{EC:3.--.--.--} (hydrolases)}.
.2 {\textbf{EC:3.2.--.--} (glycolsylases)}.
.3 {\textbf{EC:3.2.1.--} (glycosidases)}.
.4 {\textbf{EC:3.2.1.2} (hydrolysis of (1$\rightarrow$4)-\textalpha-D-glucosidic linkages in polysaccharides)}.
}
\medskip
%stopleaveout
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/ec_performance_by_top_level_class.png}%
  \caption{\textbf{ProteInfer performance for predictions within all 7 top-level enzyme groups.}}
  \label{fig:toplevel}
\end{figure}
Individual protein sequences can be annotated with zero (non-enzymatic proteins), one (enzymes with a single function) or many (multi-functional enzymes) leaf-level EC numbers. These are drawn from a total of 8,162 catalogued chemical reactions.
Our best $\texttt{F}_{max}$ was achieved by a model containing 5 residual blocks with 1100 filters each (full details in \textnormal{ Supplement}%htmlver:<a href="./latex/supplement.pdf">supplement</a>
). For the dev set, $\texttt{F}_{max}$ converged within 500,000 training steps. On the random split, the model achieves $\texttt{F}_{max}$ = 0.977 (0.976-0.978) on the held-out test data. At the corresponding confidence threshold, the model correctly predicts 96.7\% of true labels, with a false positive rate of 1.4\%. Results from the clustered test set are discussed below.
Performance was roughly similar across labels at the top of the EC hierarchy, with the highest $\texttt{F}_{max}$ score observed for ligases (0.993), and the lowest for oxidoreductases (0.963) (\Cref{fig:toplevel}A%htmlver: 
). For all classes, the precision of the network was higher than the recall at the threshold maximising $\texttt{F}_{max}$. Precision and recall can be traded off against each other by adjusting the confidence threshold at which the network outputs a prediction, creating the curves shown in \Cref{fig:toplevel}B.%htmlver:the figure below
%%%INITIALANALYSIS%%%

@@

We then assessed the ability of this method to more generally localize function within a sequence, even though the model was not trained with any explicit localization information. We selected all enzymes from Swiss-Prot that had two separate leaf-node EC labels for which our model predicted known EC labels, and these labels were mappable to corresponding Pfam labels. For each of these proteins, we obtained coarse-grained functional localization by using CAM to predict the order of the domains in the sequence and compared to the ground truth Pfam domain ordering\textnormal{(see supplement for details of the method)}%htmlver:(see <a href="./latex/supplement.pdf">supplement</a> for details of the method)
. We found that in 296 of 304 (97\%) of the cases, we correctly predicted the ordering, though we note that the set of bifunctional enzymes for which this analysis is applicable is limited in its functional diversity \textnormal{(see supplement)}%htmlver:(see <a href="./latex/supplement.pdf">supplement</a>)
. Although we did not find that fine-grained, per-residue functional localization arose from our application of CAM, we found that it reliably provided coarse-grained annotation of domains' order, as supported by Pfam. This experiment suggests that this is a promising future area for research.
\subsection*{Neural networks learn a general-purpose embedding space for protein function}
Whereas InterProScan compares each sequence against more than 50,000 %from summing integrated signatures at https://www.ebi.ac.uk/interpro/release_notes.html
individual signatures and BLAST compares against an even larger sequence database, ProteInfer uses a single deep model to extract features from sequences that directly predict protein function. One convenient property of this approach is that in the penultimate layer of the network each protein is expressed as a single point in a high-dimensional space. To investigate to what extent this space is useful in examining enzymatic function, we used the ProteInfer EC model trained on the random split to embed each test set protein sequence into a 1100-dimensional vector.

@@

There is arguably room for improvement in this regard from classical approaches. For example, the online interface to InterProScan can take ~147 seconds to process a 1500 amino acid sequence\footnote{Protein Q77Z83.}\footnote{It should be noted that the individual databases that make up InterProScan may return matches faster, with the online interface to Pfam taking 14--20 seconds for a 1500 amino acid sequence.}, while running the tool may make the search faster, doing so requires downloading a 9 GB file, with an additional 14 GB for the full set of signatures, which when installed exceeds 51 GB. Meanwhile, conducting a BLAST search against Swiss-Prot takes 34 seconds for a 1500 amino acid sequence \footnote{A target database with decreased redundancy could be built to reduce this search time, and other optimizations of BLAST have been developed.}.

An attractive property of deep learning models is that they can be run efficiently, using consumer graphics cards for acceleration. Indeed, recently, a framework has been developed to allow models developed in TensorFlow to be run locally using simply a user's browser \citep{tfjs}, but to our knowledge this has never been deployed to investigate biological sequence data. We therefore built a tool to allow near-instantaneous prediction of protein functional properties in the browser. When the user loads the tool, lightweight EC (5MB) and GO model (7MB) prediction models are downloaded and all predictions are then performed locally, with query sequences never leaving the user's computer. Inference in the browser for a 1500 amino-acid sequence takes < 1.5 seconds for both models (\Cref{sup:js_timing}%htmlver:see <a href="./latex/supplement.pdf">supplement</a>
).
% \begin{table*}[b]
% \centering
% \caption{Bla}
% \begin{tabular}{ccccc}
% \hline
% Input               & Processor & Statistic                                 & InterProScan & ProteInfer  \\ \hline
% $1\times$ 500 aa sequence  & CPU       & Inference time   & 1 min 18 secs            & x           \\ 
% $1\times$ 500 aa sequence  & GPU       & Inference time   & \xmark️          & x           \\ 
% $ (test set) \times$ 500 aa sequences & 12x CPUs       & Inference time & 12 hours, 40 minutes            & x           \\ 
% $64\times$ 500 aa sequence & GPU       & Inference time & \xmark️          & x           \\ 
% --                  & --        & Model sizes (compressed)                  & 50 GB        & 50 MB       \\ 
% --                  & --        & Deliverable within the user’s browser     & \xmark️           &\cmark \\ \hline
% \end{tabular}
% \end{table*}
% \note{
% %I tried making "chimaeric proteins" - when N terminus of amylase is added to MAPK7 that is predicted to have extracellular localisation (otherwise unchanged). With reverse, amylase functionality is unchanged but location changes to intracellular.
% }
\subsection*{Comparison to experimental data}
\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{Figures/price.png}%
  \caption{\textbf{ProteInfer predictions for a set of genes recently experimentally reannotated by high-throughput phenotyping.} ProteInfer makes confident and largely accurate predictions at the earliest levels of the EC hierarchy. Accuracy falls at the finest levels of classification (for this set of challenging genes) but fortunately the network declines to make a prediction in most cases, with every label failing to meet the threshold for positive classification. \label{fig:levels}}
\end{figure}
\begin{figure*}[b]
\centering
  \includegraphics[width=\textwidth]{Figures/tetrahydrofolate_synthase.png}
  % See https://docs.google.com/presentation/d/1PqP-Ux7far9NMYvroeZIQQ9iXi41xira7-awDC4w_ss/edit#slide=id.p for the provenance of this figure
  \caption{\textbf{Linking sequence regions to function with class activation mapping for C-1-tetrahydrofolate synthase (accession P11586).} A. Ground truth annotation of function on Uniprot \citep{uniprotlatest}. B. The three horizontal bars are the sequence region ProteInfer predicts are most involved in each corresponding reaction. This concurs with the known function localization. }
  \label{fig:cam}
\end{figure*}

@@

We note that there has been extensive work in GO label prediction evaluated on a temporally-split dataset (constructing a test set with the most recently experimentally annotated proteins), e.g., \citep{zhou2019cafa}, and stress that our comparison is based on the random and clustered splits of Swiss-Prot described above.  This approach to splitting the data into train and test has advantages and disadvantages as compared to a temporal split, which depend on the desired application for the method being evaluated.

%%%GO_METHOD%%%

We trained a single model to predict presence or absence for each of these terms and found that our network was able to achieve a precision of 0.918 and a recall of 0.854 for an F1 score of 0.885 (95\% CI: 0.882--0.887).

@@

There are multiple caveats to these comparisons. One challenge is that the completeness of Swiss-Prot's GO term annotations varies \cite{jiang2014impact}. As an extreme example, \textit{Pan paniscus} (Pygmy Chimpanzee) and \textit{Pan troglodytes} (Chimpanzee) have an identical Apolipoprotein A-II protein,\footnote{Accessions P0DM95 and Q8MIQ5.}, where the first protein has 24 GO annotations, while the latter has 143 GO annotations.\footnote{This count is done using not only the set of all labels that appear in Swiss-Prot, but also any parents of those labels.} One way this is reflected in the performance of the models is that some BLAST matches that have extremely large bit-scores are not annotated identically, and thus reduce the precision of the BLAST model. It is also important to note that our model has the advantage of being specifically trained on the UniProt labelling schema upon which it is being evaluated. InterPro works quite differently, with GO terms being assigned to families, and so inconsistencies in terms of how these are assigned can explain reduced performance -- for instance InterPro families simply do not feature all of the GO terms found in UniProt. Thus these results should be seen as specific to the task of reproducing the curated results in UniProt.

@@

%startleaveout
\section*{Bibliography}
\bibliography{../bibliography}
\onecolumn
\newpage
%stopleaveout
\include{supplement}
%startleaveout
